---
title: "Class 12: Online Lab"
author: "Austin Teel (A17293709)"
format: pdf
toc: TRUE
---


## Section 1. Proportion of G/G in a population


Downloaded CSV file from ensemble < https://www.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39730103-40060104;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel >

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl)*100
```

Now lets look at a different population. I picked the GBR.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```

Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.)/nrow(gbr)*100,2)
```


This varient that is associated with childhood asthma is more frequent in teh GBR population than in the MKL population.

Lets now dig into this further.



## Section 4: Population Scale Analysis
One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. 

So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```

> Q 13. One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.

Lets make a boxplot
```{r}
ggplot(expr) + aes(geno, exp, fill=geno)+
geom_boxplot(notch=TRUE)
```



> Q 14. You could infer that there is a higher level of expression for people with the A/A genotype when compared to the G/G genotype. The graph shows a difference in the amount of expression for each so we could infer that there may be a difference, but we can not conclude that there is a difference due to the fact that the error bars also overlap.



You could infer that there is a higher level of expression for people with the A/A genotype when compared to the G/G genotype. The graph shows a difference in the amount of expression for each so we could infer that there may be a difference, but we can not conclude that there is a difference due to the fact that the error bars also overlap.





































































