---
title: "Class 18 : Investigating Pertussis Resurgence: Mini Project"
author: "Austin Teel (A17293709)"
toc: TRUE
format: html
---

## Background 
Pertussis (a.k.a whooping cough) is a common lung infectuion caused by the bacteria B.Pertussis.

This infection can affect people all ages but is most severe for those under 1 year of age.

The CDC track the number of [reported cases in the US](https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html): 
We can "scrape" this data with the **datapasta** package.


```{r}
cdc <- data.frame(
                                 year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L,2023L,2024L,2025L),
                                cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044,7063,22538,21996)
       )
```

> Q. Make a plot of `year` vs `cases`

```{r}
library(ggplot2)
ggplot(cdc) +
  aes(year, cases) +
  geom_point()+
  geom_line()+
  labs()
```

> Q. Add some major milestones including the first wP vaccine roll-out (1946), and the switch to the newer aP vacccine (1996).

```{r}
library(ggplot2)
ggplot(cdc) +
  aes(year, cases) +
  geom_point()+
  geom_line()+
  labs()+
  geom_vline(xintercept = 1946, color = "blue", linewidth = 1, linetype = "dashed")+
  geom_vline(xintercept = 1996, color = "red", linewidth = 1, linetype = "dashed")+
  geom_vline(xintercept = 2020, color = "gray", linewidth = 1, linetype = "dashed")+
  annotate("text",
           x = 1946,
           y = max(cdc$cases),
           label = "wP",
           color = "blue",
           vjust = -0.5)+
  annotate("text",
           x = 1996,
           y = max(cdc$cases),
           label = "aP",
           color = "red",
           vjust = -0.5)+
  annotate("text",
           x = 2020,
           y = max(cdc$cases),
           label = "Covid-19 Pandemic",
           color = "gray",
           vjust = -0.5)
```

The graph shows were high case numbers in the pre 1940s, they were reduced drastically based on the rolling out of the wP vaccine. The numbers then began to rise again in the 1980s due to the fact that documentaries and studies came out that falsely linked other sicknesses to vaccines. This lead to less vaccines being given and a rise in infections.  Another correlation that can be made from the graph below and requires further research is the fact that the aP induced protection wanes faster than wP but why? and does this cause the rise of infection as well?



**Why is this vaccine-preventable disease on the upswing?** To answer this question we need to investigate the mechanisms underlying waning protection against pertussis. This requires evaluation of pertussis-specific immune responses over time in wP and aP vaccinated individuals.

# CMI-PB project
[Computational Models of Immunity](https://www.cmi-pb.org/) - pertussis boost project aims to provide the scientific community with this very information.

They make their data available via JSON format returning API. We can read this in R with the `read_json()` function from the **jsonlite** package: 

```{r}
library(jsonlite)

subject <- read_json("http://cmi-pb.org/api/v5_1/subject", simplifyVector=TRUE)

head(subject)
```


> Q. How many wP and aP individuals are in this `subject` table?


```{r}
table(subject$infancy_vac)
```

> Q. What is the biological sex breakdown?

```{r}
table(subject$biological_sex)
```

> Q. In terms of race and gender is this dataset representative of the US population?

```{r}
table(subject$race, subject$biological_sex)
```

Let's read some more database tables:

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE)
ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE)
```

```{r}
head(ab_titer)
```

To analyze this data we need to first "join" (merge) the different tables so we have all the data in one place not spread accross different tables.

We can use the `*_join()` family of functions from **dplyr** to do this

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
head(meta)
```

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```

> Q. What antibody isotypes are measure for these patients?

```{r}
table(abdata$isotype)
```

> Q. What antigens are reported?

```{r}
table(abdata$antigen)
```

> Q. How many datasets are there in the abdata?

```{r}
unique(abdata$dataset)
```

>Q. How many times do the datasets appear in the abdata? what do you notice from this?

```{r}
table (abdata$dataset)
```

We can see 4 different data sets and from them we can see that there were so many data sets in 2020 likely the result of the covid pandemic while there was an extremely low amount in the 2021 and 2022 data sets with an increase again in 2023. Although something to notice is tat even in 2023 we dont have half of the entries that we did from 2020.


Let's focus on the IgG isotype and make a plot of MFI_normalized for all antigens.

```{r}
igg <- abdata |>
  filter(isotype == "IgG")
head(igg)
```

```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen)+
  geom_boxplot()
```

> Q. Is there a difference for aP vs wP individuals with these values?

```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen)+
  geom_boxplot()+
  facet_wrap(~infancy_vac)
```

```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen, col=infancy_vac)+
  geom_boxplot()
```

> Q. Is there a temprol response - i.e. do values increase or decrease over time?

```{r}
ggplot(igg)+
  aes(MFI_normalised, antigen, col=infancy_vac)+
  geom_boxplot()+
  facet_wrap(~visit)
```

## FOcus on "PT" Pertussis TOxin antigen

```{r}
pt.igg.21 <- igg |> filter(antigen == "PT", 
              dataset == "2021_dataset")
```

```{r}
ggplot(pt.igg.21)+
  aes(planned_day_relative_to_boost, 
      MFI_normalised, 
      col=infancy_vac,
      group=subject_id)+
  geom_point()+
  geom_line()+
    geom_vline(xintercept = 14, lty=12)
```

```{r, warning=FALSE}
ggplot(pt.igg.21) +
  aes(planned_day_relative_to_boost,
      MFI_normalised,
      col = infancy_vac) +

 
  geom_line(aes(group = subject_id),
            alpha = 0.3,      
            linewidth = 0.6) +

  geom_point(alpha = 0.3) +


  geom_smooth(method = "loess",
              span=0.4,
              se = FALSE,
              linewidth = 1.2,
              alpha = 1) +    

  geom_vline(xintercept = 14, lty = 12)
```






































































































































































