---
title: "class06extra"
author: Austin Teel (A17293709)
format: pdf
---


## (B. Can you improve this analysis code?

```{r}
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s1, chain="A", elety="CA")
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```


We must first normalize a function for the data in order to smoothly apply it to all of the values. This function will take the PDB id values. Then it will get a proetin structure from the pdb which it tehen can find the mapped points of the B-factor values. This then will give us the numeric vector of the B-factor values of the protein.
```{r}
library(bio3d)
#The input of the function is the pbd id values such as 4AKE,1AKE, and 1E4Y.
process_and_plot <- function(pdb_id) {
  pdb <- read.pdb(pdb_id)
#The following code will allow the function to focus on the alpha-carbon atoms (CA).
#Then it will also focus on the b-values.
  chainA <- trim.pdb(pdb, chain = "A", elety = "CA")
  bvals <- chainA$atom$b
#The following piece of the function will graph the values along the sequence
  plotb3(bvals, sse = chainA, typ = "l", ylab = paste("Bfactor:", pdb_id))
#Then will return numeric vectors of the b-values.
  return(bvals)
}
```

Then we must apply the function above to the data. We can do this simply by calling the function that we wrote above and applying it to the values b1-b3.

```{r}
b1 <- process_and_plot("4AKE")
b2 <- process_and_plot("1AKE")
b3 <- process_and_plot("1E4Y")
```

We can see that the resulting graphs are comparable to the initial graphs and represent the same proteins based off of their structure based off of there pdb id. In this case we are just using a normalized function.