---
title: "Class06: R functions"
author: "Austin Teel (A17293709)"
format: pdf
toc: TRUE
---

## Background

Functions are at the heart of using R. Everything we do involves calling and using functions (from data input, analysis to results output).

All functions in R have at least 3 things:

1. A **name** the thing we use to call the function.
2. One or more input **arguments** that are comma seperated
3. The **body**, lines of code between curly brackets {} that does the work of the function.

## A first function

Let's write a silly wee function to add some numbers: 

```{r}
add <- function(x) {
  x+1
}
```

Lets try it out
```{r}
add(100)
```

Will this work

```{r}
add( c(100,200,300))
```
Modify to be more useful and add more than just 1

```{r}
add <- function(x,y=1) {
  x+y
}
```


```{r}
add(100,10)
```
Will this still work
```{r}
add(100)
```

```{r}
plot(1:10, col="blue", typ="b")
```

```{r}
log(10, base=10)
```

> **N.B.** input arguments can be either **required** or **optional**. The later have a fall-back default that is specified in the function code with an equals sign.


```{r}
#add(x=100,y=200,z=300)
```


## A second function

All function sin R look like this 

```
name <- function(arg) {
body
}
```

The `sample()` function in R takes random samples from the group of numbers it represents it can be with or without replacement based on the arguments you give it. It defaults to picking without replacement. Only need sample amounts and size, the rest is optional.

```{r}
sample(1:10,size=4)
```

> Q. Return 12 numbers picked randomly from the input 1:10

```{r}
sample(1:10,size=12, replace=TRUE)
```


>Q. Write the code to generate a random 12 nucleotide long DNA sequence.

```{r}
bases <- c("A","C","G","T")
sample(bases, size=12, replace=TRUE)
```

> Q. Write a first version function called `Generate_dna()` that generates a user specified length `n` random DNA sequence?

```{r}
name <- function(arg){
  
}
```


```{r}
generate_dna <- function(n=6) {
  sample(bases, size=n, replace=TRUE)
}
```

```{r}
generate_dna(100)
```

> Q. Modify you function ot return a FASTA like sequence so rather than [1] "T" "A" "G" "T" "C" "G" "C" we want "TAGTCGC"

```{r}
generate_dna <- function(n=6) {
  paste(sample(bases, size=n, replace=TRUE),collapse="")
}
```


```{r}
generate_dna(100)
```


> Q. Give the user an option to return FASTA format output sequence or standard multi-element vector format?

```{r}
generate_dna <- function(n=6, fasta= TRUE) {
  ans <- sample(bases, size=n, replace=TRUE)
  
  if(fasta) {ans <- paste(ans,collapse = "")}
  
return(ans)  

}
```

```{r}
generate_dna(20)
generate_dna(20,FALSE)
```

##A new cool function

> Q. Write a function called `generate_protein()` that generates a user specified length protein sequence in FASTA like format?

```{r}
aa <- c("A","R","N","D","C","E","Q","G","H","I",
  "L","K","M","F","P","S","T","W","Y","V")

generate_protein <- function(n=6, fasta= TRUE) {


aas <- sample(aa, size=n, replace=TRUE)

  
  if(fasta) {paste(aas,collapse = "")}
return(paste(aas,collapse = ""))
}

generate_protein(10)
```

> Q. Use your new `generate_protein()` function to generate sequences between length 6 and 12 amino acids in length and check if any of these are unique in nature (i.e. found in the NR database at NCBI)?

```{r}
aa <- c("A","R","N","D","C","E","Q","G","H","I",
  "L","K","M","F","P","S","T","W","Y","V")

generate_protein <- function(n=6, fasta= TRUE) {


aas <- sample(aa, size=n, replace=TRUE)

  
  if(fasta) {paste(aas,collapse = "")}
return(paste(aas,collapse = ""))
}

generate_protein(6)
generate_protein(7)
generate_protein(8)
generate_protein(9)
generate_protein(10)
generate_protein(11)
generate_protein(12)
```

Or we could do a `for()` loop:

```{r}
for(i in 6:12){
  cat(">", i, sep="", "\n")
  cat(generate_protein(i),"\n")
}
```




